Draft

### 2019-06-25 draft.tex
Proved that 2 is B-null in extensional TT.

\usepackage{setspace}
%\usepackage{graphicx}
\usepackage{enumitem}
%\usepackage{tikz-cd}
%\usepackage{todonotes}
\usepackage{multicol}
\usepackage{mathrsfs}
\usepackage[margin=1in]{geometry} 
\usepackage{amsmath,amsthm,amssymb}
\usepackage{ textcomp }
\usepackage{ wasysym }
\usepackage{bookmark}
\usepackage{accents}
\usepackage[parfill]{parskip}

@@

\section{unassigned}

\begin{defn}
The Independence Principle for $X : \mathcal{U}$ and family of propositions $Q: X \rightarrow N \rightarrow \mathtt{Prop}$ is the following statement
$$\mathtt{IP}_{X,Q} : \Big(\prod_{x : X}\| \Sigma_{n : \mathbb{N}} Q(x,n) \| \Big) \rightarrow \| \Sigma_{n : \mathbb{N}}\prod_{x : X} Q(x,n) \|$$
\end{defn}
We know that there exists model of extensional type theory that validates IP. 
We would like a variant of IP to which we can apply Theorem 6.1 of CTCA. 
We can then reach the desired form by omitting the truncation in the antecedent of IP. 
\begin{defn}
We define $\mathtt{IP'}_{X,Q}$ with $X : \mathcal{U}$ and family of propositions $Q: X \rightarrow N \rightarrow \mathtt{Prop}$ 
$$\mathtt{IP}_{X,Q} : \Big(\prod_{x : X} \Sigma_{n : \mathbb{N}} Q(x,n)  \Big) \rightarrow \| \Sigma_{n : \mathbb{N}}\prod_{x : X} Q(x,n) \|$$
\end{defn}
Firstly, $\mathtt{IP}_{X,Q} \Rightarrow \mathtt{IP'}_{X,Q}$. 
Secondly, $\mathtt{IP'}$ is strong enough to prove that $\mathbb{N}$ is $B$-null, where 
$$B : A \rightarrow \mathcal{U}$$
$$S :\equiv \mathbb{N} \rightarrow \mathbf{2}$$
$$A : \Sigma_{s : S}\mathtt({atMost1one}\; s)$$
$$B : \prod_{a : A}\|\mathtt{p_{odd}}\; a.\mathtt{fst} + \mathtt{p_{even}}\; a.\mathtt{fst}\|$$
Specifically, it is enough to make sure that 
$$\prod_{a : A}\prod_{f : B'(a) \rightarrow \mathbb{N}} \mathtt{IP}_{B'(a),Q(a, f)}$$
where $B'(a)$ is the untruncation of $B(a)$ and
$$Q(a, f) : \prod_{b : B'(a)} \prod_{n : \mathbb{N}} \mathcal{U}$$
and is defined as 
$$Q(a, f)\; b\; n \equiv f(b) =_{\mathbb{N}} n$$

>>>>

\usepackage{setspace}
%\usepackage{graphicx}
\usepackage{enumitem}
%\usepackage{tikz-cd}
%\usepackage{todonotes}
%\usepackage{multicol}
%\usepackage{mathrsfs}
%\usepackage[margin=1in]{geometry} 
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{ textcomp }
%\usepackage{ wasysym }
\usepackage{bookmark}
%\usepackage{accents}
\usepackage[parfill]{parskip}

@@

\section{Extensional Type Theory}
There exists a model of extensional type theory that validates the following `Independence Principle'
$$\bigg( \prod_{s : \mathbf{2}^\mathbb{N}} P\; s \rightarrow \Big(\prod_{n : \mathbb{N}}s\; n = 0 \Big) \rightarrow \Big\lVert \sum_{z : \mathbf{2}} Q\; s\; z \Big\rVert \bigg)$$
$$\rightarrow \bigg( \prod_{s : \mathbf{2}^\mathbb{N}} P\; s \rightarrow  \Big\lVert \sum_{z : \mathbf{2}} \Big(\prod_{n : \mathbb{N}}s\; n = 0 \Big) \rightarrow Q \; s\; z\Big\rVert \bigg)$$
where $P: \mathbf{2}^\mathbb{N} \rightarrow \mathtt{Prop}$ and $Q : \mathbf{2}^\mathbb{N} \rightarrow \mathbf{2} \rightarrow \mathtt{Prop} $ are families of propositions. 
Putting it plainly, if the left hand side of the above is true, then $z : \mathbf{2}$ does not depend on the proof of the constantness of $s$ to $0$.

The same model also validates Markov's Principle
$$\Big(\prod_{a : A}(P\; a \vee \neg P\; a)\Big) \rightarrow  \Big(\neg \prod_{a : A}\neg P\; a\Big)\rightarrow \Big\lVert \sum_{a : A} P\; a \Big\rVert$$
where $P : A \rightarrow \mathtt{Prop}$ is a family of propositions over $A : \mathcal{U}$. 
Informally, Markov's Principle states that if we have a collection of decidable propositions and the fact that not all of them are false, then one of them must be true.

In this section we will be working in the context of this model.

Let $A :\equiv \sum_{a: \mathbf{2}^\mathbb{N}}\mathtt{atMost1one}\;a$. 
When referring to elements of $A$ we implicitly mean the first part of the pair. 
Let
$$B : A \rightarrow \mathtt{Prop}$$
$$B :\equiv \lambda\; a.\; \mathtt{p_{odd}}\; a \vee \mathtt{p_{even}}\; a$$
\begin{lem}
$\mathbf{2}$ is $B$-null.
\end{lem}
\begin{proof}
Given $a : A$ and $f : B\; a \rightarrow \mathbf{2}$ we need to prove that there exists unique $f' : \mathbf{1} \rightarrow \mathbf{2}$ through which $f$ factors, in the sense that $f = f' \circ g$, where $g : B\; a \rightarrow \mathbf{1}$ is the sole inhabitant of its function space. 
Functions with $\mathbf{1}$ as their domain are constant. 
We therefore need to find some $z : \mathbf{2}$ so that $\lambda\; \_.\; z$ is $f'$. 
To that end, we will use the following instance of the Independence Principle
$$\bigg( \prod_{s : \mathbf{2}^\mathbb{N}} \Big(\big(\prod_{n : \mathbb{N}} s(n) = a(2 \cdot n)\big) \vee \big(\prod_{n : \mathbb{N}} s(n) = a(2\cdot n +1)\big) \Big) \rightarrow \Big(\prod_{n : \mathbb{N}}s\; n = 0 \Big) \rightarrow \Big\lVert \sum_{z : \mathbf{2}}\prod_{p : B\; a} f\; p = z\Big\rVert \bigg)$$
$$\rightarrow \bigg( \prod_{s : \mathbf{2}^\mathbb{N}} \Big(\big(\prod_{n : \mathbb{N}} s(n) = a(2 \cdot n)\big) \vee \big(\prod_{n : \mathbb{N}} s(n) = a(2\cdot n +1)\big) \Big) \rightarrow  \Big\lVert \sum_{z : \mathbf{2}} \Big(\prod_{n : \mathbb{N}}s\; n = 0 \Big) \rightarrow\prod_{p : B\; a} f\; p = z\Big\rVert \bigg)$$
Note that the left hand side is provable. 
By providing the odd $s_1$ and even $s_2$ subsequences of $a$ to the right hand side, we get hold of $z_1 : \mathbf{2}$ and $z_2 : \mathbf{2} $ respectively. 
Equality on $\mathbf{2}$ is decidable, therefore $z_1 = z_2 \vee z_1 \neq z_2$ is provable. 
Since our goal is a proposition, we can drop the truncation of the disjunction. 
This leaves us with a coproduct.

We will prove that the fact $\mathbf{2}$ is $B$-null follows from both constituents of the coproduct.
\begin{itemize}
\item First the case where $z_1= z_2$. 
We arbitrarily pick $z_1$ and propose $f' : \equiv \lambda \;\_.\; z_1$. 
By axiom K we have function extensionality, so we reduce proving $f= f' \circ g$ to proving $f\; b = f'\circ g\; b$ for arbitrary $b : B\; a$. \\
$b$ is a truncated coproduct witness. 
As before we can ignore the truncation and take cases on the coproduct. 
\begin{itemize}
\item We first consider the case $\mathtt{p_{odd}}\; a$. 
This tells us that the odd subsequence is constantly $0$, which we can provide to the rhs of the IP alongside with $b$  itself (and $s_1$ as the first argument) to get $f\; = z_1$. We concatenate this with $f' \circ g \; b = z_1$ to conclude this case. 
\item The $\mathtt{p_{even}}$ case is very similar. 
The only difference is that we provide $s_2$ as the first argument to the rhs of the IP and then have $z_1=z_2$ be part of the concatenation of paths when we prove homotopy between $f$ and $f' \circ g$. 
\end{itemize}

\item Now consider the case $z_1 \neq z_2$. 
For any $n : \mathbb{N}$, $a\; n = 1$ is decidable. 
Suppose that $\prod_{n : 
\mathbb{N}} a\; n \neq 1$. 
Then $a$ and by extension $s_1$ and $s_2$ are constantly $0$. 
By the rhs of the IP we then have $z_1 = f\; b = z_2$ where $b : B\; a$ is clearly available to us at this point. 
This constradicts $z_1 \neq z_2$. 
We have proven $\neg \prod_{n : 
\mathbb{N}} a\; n \neq 1$. 
By Markov's Principle there exists $n_1 : \mathbb{N}$ such that $a\; n_1 = 1$. 
We pick the subsequence $s_0$ with parity opposite to $n_1$ and let $f'$ be always equal to the corresponding $z_0$. 
We prove $\prod_{n : \mathbb{N}} s_0\; n = 0$ by induction on $\mathbb{N}$ and cases on $s_0 \;n$. 
In the case where $s_0\; n = 1$ we can reach falsum because $n$ will have to be both even and odd (in $a$) by $\mathtt{atMost1one} \; a$. Ex falso trivializes this case. 
Now that we have established this too, we have enough arguments for the rhs of IP to output the desired equality $f\; b = z_0$ which proves homotopy between $f$ and $(\lambda\;\_.\; z_0) \circ g$.
\end{itemize}
\end{proof}
